Modelling the emergence of whisker barrels

### 2020-07-02 paper/briefpaper/brief.tex
Brings in updated text from Stu's google doc

First we verified that all results established by \cite{karbowski_model_2004}
for a 1D axis could be reproduced using our extension to a 2D cortical
sheet. Using an elliptical domain, $S$, with $M=3$ offset guidance gradients
aligned to the longer axis, $N=5$ thalamocortical projections gave rise to
five distinct cortical fields at locations that preserved the topographic
ordering defined by the original $\gamma$ values. However, we found that
specifying $N$ ordered areas required $M\approx (N+1)/2$ signalling
fields. This is because localization of axon densities occurs only when
projections are influenced by interactions with two or more signalling
gradients that encourage migration in opposing directions.
%
%Note that these dynamics are quite unlike classic chemospecificity models
%\citep{sperry_chemoaffinity_1963}, which essentially assume center-points
%i.e., conditions in the target tissue that instruct pre-identified afferents
%to stop growing.
%
As the number of distinct guidance fields is unlikely to
approach the number of barrels, \cmnt{modifications to the model} were required.

@@

\mpfive{Testing this reasoning required increasing the strength of the
  competition between simulated thalamocortical projections for cortical
  territory, by increasing the tendency for the density of each projection to
  concentrate. The major modification required was thus to introduce into the
  model an additional source of competition between thalamic projections.}
%
The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching,
\MPtwo{whilst ensuring that $a_i$ is conserved over time,} we
redefined
%
\begin{equation} \label{eq:comp}
  \color{revblue}
  \chi_i(\mb{x}, t) = \frac{\epsilon a_i}{N-1} \nabla \textstyle\sum_{j\ne i}^{N}{a_j}.
  \color{black}
\end{equation}
%
\mpfive{This term contributes to the \emph{flux} of axonal branching as
  an additional source of diffusion, which reduces the branching density for a
  given projection where the branches of other projections are dense. The
  strength of this additional competition is parameterized by
  $\epsilon$}. Note that this operation is local to individual afferent
projections.

\MPone{In addition, the model outlined requires that molecular
  guidance gradients in the cortex are complemented by graded values of the
  interaction strengths, $\gamma$, at the level of the thalamus. While the
  precise mechanisms by which thalamic and cortical gradients interact during
  development have not been fully characterised, the presence of complementary
  thalamic and cortical molecular guidance gradients has been well established
  experimentally. In particular, the EphA4 receptor and its ligand ephrin-A5
  are distributed in complementary gradients in the somatosensory thalamus and
  cortex} (\citealp{vanderhaeghen_mapping_2000,miller_epha7-ephrin-a5_2006}).
\MPone{Cells originating in the medial part of ventrobasal thalamus
  (comprising VPM and VPL) express high levels of EphA receptors and project
  to the lateral part of S1, which expresses low levels of ephrin-A5, and
  cells originating in the lateral part of ventrobasal thalamus express low
  levels of EphA receptors and project to the medial part of S1, which
  expresses high levels of ephrin-A5} (see
\citealp{gao_regulation_1998,dufour_area_2003,vanderhaeghen_developmental_2004,speer_grading_2005,torii_role_2013}).

\MPone{For simplicity, both the thalamic interaction gradients and the
  cortical guidance fields in the model were chosen in the first instance to
  be linear and orthogonal. Hence a given pair of gamma values corresponds to
  the coordinate of a barreloid center in the VPM (in a reference plane
  defined by the anterior-posterior and medial-lateral axes. Coordinates were
  estimated from Fig 5d of} \citep{haidarliu_size_2001}, \MPone{and
  scaled (arbitrarily) such that $\gamma\in\pm 2$.}

A cortical boundary enclosing \cmnt{the} 41 \cmnt{macrovibrissae} barrels was
traced from a cytochrome oxidase stain from \cite{zheng_signal_2001}
\cmnt{(using original data kindly supplied by the authors)}, and
Eqs.~\ref{eq:dc}--\ref{eq:comp} were solved for $N=41$ projections on the
resulting domain, using $M=2$ linear signalling gradients aligned with the
anterior-posterior and medial-lateral axes.  \mpfour{These gradients may
  extend well beyond the boundary of the barrel field (e.g. as ephrin-5 does;
  see \citealp{miller_epha7-ephrin-a5_2006}) but they are shown only within
  the boundary in Fig.\,\ref{fig:main}A, to indicate that external values are
  not used by the simulation. Simulations were stepped through 30000
  iterations of Eqs.\,\ref{eq:dc}--\ref{eq:comp} ($\delta t=0.0001$)}

From initial conditions sample from a uniform random distribution for
$a(\mb{x},0)\in(0.2,0.4)$ and $c(\mb{x},0)=0$, and across a wide range of
parameter values,
%From random initial conditions for $a(\mb{x},0)$, \MPtwo{with each hexagonal element
%  of $a_i$ set from a uniform distribution according to $0.2+{U}(0,0.2)$, and
%  $c(\mb{x},0)=0$},
a clear Voronoi-like tesselation \cmnt{of topographically organised
  projections} emerged, \cmnt{confirming that barrel maps can self-organise in
  the absence of pre-specified center points. Parameters for the example
  simulation shown in Fig.\,\ref{fig:main}C (see Movie S1) were obtained by
  conducting a full parameter sweep and choosing a combination of values that
  scored well against three measures.}

% New section to describe and introduce the metrics
\metrics{First, we used an algorithm introduced by Honda to measure the
  discrepancy of each barrel shape from a Dirichletform shape. Low overall
  values of the dimensionless \emph{Honda-}$\delta$ metric (called $\Delta$
  in}
\citealp{honda_geometrical_1983}) \metrics{obtained from simulated barrels indicate a close
  correspondence of the simulated barrel field with a Voronoi tesselation, and
  thus with a biological barrel field. For the
  % Run ./build/sim/honda boundaries/rat_barrels/wb_110405_Dirichlet.svg to
  % get this result:
  tesselation overlaid on the real barrel field in Fig.\,\ref{fig:main}A,
  $\delta=0.025$, and a reduction in the Honda-$\delta$ metric in simulation
  confirmed that an equivalent `good' Voronoi pattern} (see
\citealp{senft_mouse_1991}) \metrics{emerged within $\approx$~20000 iterations
  (Fig.\,\ref{fig:main}D, red circles). Second, we devised a \emph{pattern
    difference} measure that is sensitive to deviations in the component
  shapes and overall topographic registration of two tesselations, $\eta$, and
  we used this measure to compare the simulated barrel fields to the real
  barrel field from which the simulation boundary shape was obtained (see
  Materials and Methods for details). A similar reduction in pattern
  difference over simulation time confirmed that the shapes and arrangement of
  emergent connection fields matched those of the real barrel field by around
  20000 iterations. Third, we measured the \emph{connection selectivity} at
  each location on the cortical sheet, as the connection density of the most
  dense projection divided by the sum over all projection densities,
  $\omega$.}
%
\metrics{Against these three metrics we are also able to characterise the
  robustness of self-organization to the model parameters, and to investigate
  the sensitivity of the model to variation in its inputs.}

% Most of this commented out region is covered in the para above.
%
%A reduction in the Honda delta
%metric (see \citealp{senft_mouse_1991}) confirmed that a `good' Voronoi
%pattern emerged within \cmnt{$\approx$~20000} iterations
%(Fig.\,\ref{fig:main}D).  A measure of the difference in \cmnt{barrel area}
%between real and simulated barrels revealed a strong correspondence (see
%Fig.\,\ref{fig:main}D, squares). \cmnt{In a chosen region, the ratio of
%  \emph{the connection density of the TC type $i$ which makes the densest
%    connections} to \emph{the sum of all connection densities} gives a measure
%  of the selectivity of the connection density, $\mu(\mb{x},t)$. We summed this measure
%  across the entire barrel domain and named it $\omega(t)$ (see
%  Fig.\,\ref{fig:main}D, grey hexagons). $\omega$ increases during the first
%  20000 simulation steps and converges on a stable
%  value. Fig.\,\ref{fig:main}E shows the selectivity measure plotted as a
%  luminance map, giving a more graded impression of the structure which
%  develops in the system.}

@@

\MPthreePar{To determine the robustness of the model to its parameters, we
  investigated its behaviour with respect to changes in the parameters $D$,
  $\epsilon$ and the ratio of $\alpha$ to $\beta$. Fig.\,\ref{fig:paramsweep}A
  shows the value of the metrics $\delta$, $\eta$ and $\omega$ after 30000
  iterations, plotted as colour maps, for 216 parameter sets. The maps
  indicate that the pattern is not strongly influenced by the choice of
  $\alpha/\beta$, which was varied by 5 orders of magnitude. The most
  Voronoi-like patterns were achieved for parameter sets for which $D$ and
  $\epsilon$ were high. The barrel area metric, $\eta$ indicates a strong
  preference for $D\approx0.5$, although this does not give the solution of
  maximum selectivity, which favours lower values of $D$. Thus, there is a
  trade-off between good selection and good Voroni pattern conformation.  The
  parameter set indicated by the asterisk in Fig.\,\ref{fig:paramsweep}A was
  chosen for Fig.\,\ref{fig:main} because it provides good pattern
  correspondence and a close approximation to a Voronoi tesselation.}

\MPthreePar{In Fig.\,\ref{fig:paramsweep}B, we present some patterns
  representing alternative parameter choices. Because the patterns are robust
  to alterations in $\epsilon$ and the choices of $\alpha$ and $\beta$, we
  concentrate our attention on the diffusion constant, $D$. Changing $D$,
  whose value defines the rate at which a region of localized branching will
  spread out across the domain, $S$, is equivalent to making changes to the
  size of $S$ with a fixed branching diffusion rate. These simulations can
  therefore be thought of as representing animals with a varying cortical
  areas but with unchanged cortical complexity. The most extreme example
  (Fig.\,\ref{fig:paramsweep}B iii) could represent an animal with a very
  large cortical area (such as the South American capybara) which receives
  sensory input from $N=41$ thalamic barreloids. Here, a large proportion of
  the domain is occupied by the projections with intermediate interaction
  parameters and those with strong interaction parameters are bunched up
  towards the edge of the domain. In this case, a barrel pattern fails to
  form, agreeing with the observation that cortical whisker barrels are absent
  in the capybara.}

@@

% This section needs to emphasize that the absolute values of gamma are not important
\MPthreeSens{We performed a sensitivity analysis to determine if the quality
  of the pattern (i.e.~the values of $\delta$, $\eta$ \& $\omega$ after
  $t=30000$ steps) was affected by perturbations to any of the following
  inputs:
%
  i)~the magnitude and offset of the noise applied to $a_i$ at $t=0$;
%
  ii)~random noise applied to the interaction parameters, $\gamma_{i,j}$;
%
  iii)~random noise (at various length scales) applied to the guidance fields;
%
  and
%
  iv)~The angle and magnitude of one of the guidance fields, ($\rho_1$;
  Fig.\,\ref{fig:main}A).
}

\MPthreeSens{First, we conducted 10 runs of the simulation shown in
  Fig.\,\ref{fig:main}C to evaluate the mean and standard deviation of the
  metrics in the default model. The results (where deviations occur as a
  result of the different random initial conditions for $a_i$) were
  $\delta=0.089\,(0.004)$, $\eta=0.2108\,(0.002)$ and
  $\omega=0.2165\,(0.0001)$. The pattern was barely affected by the random
  changes to $a_i(t=0)$ on each run.}
%
\MPthreeSens{Next, we modified the parameters of the generating uniform
  distribution to initialize $a_i$ for each hex to $0.1+{U}(0,0.4)$,
  increasing the variability of $a_i(\mb{x},t=0)$ without changing the mean
  initial activation. We ran this system 10 times and obtained the results
  $\delta=0.090\,(0.005)$, $\eta=0.2136\,(0.004)$ and
  $\omega=0.2166\,(0.0003)$. Paired two-sample t-tests demonstrated that the
  increase in noise amplitude made no significant difference to the metric
  results ($\delta: \mathrm{t}=-0.69,\mathrm{p}=0.49$, $\eta:
  \mathrm{t}=-2.0,\mathrm{p}=0.06$, $\omega:
  \mathrm{t}=-1.7,\mathrm{p}=0.11$). 10 runs of a noise-free initial condition
  with $a_i=0.3$ for every hex showed that the model is deterministic (all
  runs produced the results $\delta=0.0928$, $\eta=0.21236$ and
  $\omega=0.21649$) and that stochasticity in $a_i(\mb{x},t=0)$ is not
  required by this model, as all three results lay within one standard
  deviation of those for the runs in which $a_i$ was initialized
  stochastically.}

% Quick check that noise free result is within 1 SD for the noisier initial conditions
%         Noisefree    Normal noise    NF SD from Normal?   Noisy            NF SD from Noisy?
% delta   0.0928       0.089(0.004)    <1                   0.090 (0.005)    <1
% eta     0.2124       0.2108(0.002)   <1                   0.2136 (0.004)   <1
% omega   0.21649      0.2165(0.0001)  <1                   0.2166 (0.0003)  <1

\MPthreeSens{Fig\,\ref{fig:sens}A shows the results for perturbations ii) to
  iv). The results show that the Voronoi nature of the pattern ($\delta$) and
  the selectivity of the fields are not strongly affected by interaction
  parameter noise, but the pattern metric, $\eta$ shows a degradation in the
  pattern arrangement when a) noise in the interaction parameters causes their
  disordering (find better way to express?), b) noise in the guidance field
  with a high magnitude and short length scale causes disruption to the barrel
  edges and c) the guidance fields tend towards a non-orthogonal arrangement,
  or if one field becomes substantially stronger in its interaction with the
  thalamocortical projections. Fig\,\ref{fig:sens}B shows perturbed examples.}

% This should be a little less terse than in the original.
\MPthreePred{To further investigate the interplay of genetic and intrinsic
  mechanisms we simulated two well known experimental manipulations of the
  whisker barrel system. First, we simulated a seminal barrel duplication
  experiments}
\citep{shimogori_fibroblast_2005,assimacopoulos_fibroblast_2012}
\MPthreePred{in which the growth factor Fgf8, which is normally expressed at
  the anterior end of the cortical sub-plate from around E9.5}
\citep{crossley_mouse_1995}, \MPthreePred{is ectopically expressed (by
  electroporation) also at the posterior end. Because Fgf8 sets off a chain of
  gene interactions which are involved in the control of the expression of
  ephrin and cadherin molecules in gradients across the cortical plate, its
  misexpression at the posterior pole of a cortical hemisphere could be
  expected to mirror those gradients which usually vary along the
  rostro-caudal axis. We assume that this results in a mirror of the primary
  barrel cortex boundary along the rostro-caudal axis, to match the
  experimental results of}
\citealp{assimacopoulos_fibroblast_2012}. \MPthreePred{We also assume that a
  further consequence of the posterior Fgf8 electroporation is the mirrored
  anterior-posterior guidance gradient $\rho_1$ (Fig.\,\ref{fig:fgf8}A). We
  mirrored this across a reflection line which is orthogonal to its angle
  $\phi_{\rho_1}=6$\textdegree, and left $\rho_2$ unmodified. Although the
  rostro-caudal boundary reflection is not perfectly aligned with $\rho_1$,
  and we might expect some modification of both $\rho_1$ \emph{and} $\rho_2$,
  mirroring only $\rho_1$ retains the simplicity of the guidance fields; they
  remain either linear ($\rho_2$), or piece-wise linear ($\rho_1$).}

\MPthreePred{The new boundary shape and $\rho_1$ guidance gradient were the
  only changes made to the simulation of Fig.\,\ref{fig:main}; the simulation
  was run with $N=41$ thalamocortical axon types as before and the initial
  values for $a_i$ were randomly initialised according to the same
  distribution. The result, after 30000 simulation steps, was two
  mirror-symmetrical barrel fields comprising $2N$ barrels
  (Fig\,\ref{fig:main}B). This result suggests that by the misexpression of
  Fgf8,} \cite{shimogori_fibroblast_2005} and \cite{assimacopoulos_fibroblast_2012}
  \MPthreePred{created the correct cortical
  conditions for the self-organization of the two mirrored barrel fields. }

%However, merging the two fields to
%create an extended mirror-symmetric boundary shape, with an anterior-posterior
%guidance field reflected at the join, gave rise to two mirror-symmetrical
%barrel fields comprising 2N barrels, i.e., two cortical fields for each
%thalamic projection (Fig.\,\ref{fig:fgf8}). Together these results suggest that by the
%misexpression of Fgf8, \cite{shimogori_fibroblast_2005}
%effectively created one large barrel field rather than two distinct cortical
%fields.

@@

\MPthreePred{To investigate the response of the model to environmental
  manipulation, we simulated a whisker deprivation experiment. In the first
  few days after birth, the upper region of the rat's cortex develops into
  distinct layers I--IV (layers V and VI are already distinct at
  birth). During this critical period, removal of the mystacial whiskers by
  electrocauterization, plucking or trimming leads to observable changes in
  brain structures, including the barrel field in cortical layer
  IV~\citep{jeanmonod_mouse_1981}. Amoung other changes, barrels associated
  with deprived whiskers have a smaller area. This is found whether the animal
  is deprived of an individual whisker~\citep{kossut_effects_1992}, or an
  entire row~\citep{loos_somatosensory_1973}, and demonstrates that
  thalamocortical activity contributes to the developmental process. We
  simulated the trimming of the individual whisker C3 during this critical
  period by reducing the amount by which its thalamocortical projections
  compete for branching. This was achieved by reducing the value of $\epsilon$
  in Eq.\,\ref{eq:comp} for the C3 projection, leaving other simulation
  parameters unchanged. The reduced competition for branching limited the area
  over which the C3 projection could make connections and its resulting barrel
  size was diminished (Fig.\,\ref{fig:fgf8}C). Its neighbours, whose
  ability to compete for branching was unchanged, were able to increase in
  size compared to the control simulation presented in Fig\,\ref{fig:main}C
  (Fig.\,\ref{fig:fgf8}D). The C3 barrel disappeared altogether if its
  competition parameter was reduced below a half of the original value.}

@@

>>>>

First we verified that all results established by \cite{karbowski_model_2004}
for a 1D axis could be reproduced using our extension to a 2D cortical
sheet. Using an elliptical domain, $S$, with $M=3$ offset guidance gradients
aligned to the longer axis, $N=5$ thalamocortical projections gave rise to
five distinct cortical fields at locations that preserved the topographic
ordering defined by the original $\gamma$ values. However, we found that
specifying $N$ ordered areas required $M\approx (N+1)/2$ signalling
fields. This is because localization of axon densities occurs only when
projections are influenced by interactions with two or more signalling
gradients that encourage migration in opposing directions. As the number of
guidance fields is unlikely to approach the number of individual barrels,
\cmnt{modifications to the model} were required.

@@

\mpfive{Testing this reasoning required increasing the strength of the
  competition between simulated thalamocortical projections for cortical
  territory, by increasing the tendency for each projection to compete for
  cortical space into which to concentrate. The major modification required
  was thus to introduce into the model an additional source of competition
  between thalamic projections.}
%
The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching,
\MPtwo{whilst ensuring that $a_i$ is conserved over time,} we
redefined
%
\begin{equation} \label{eq:comp}
  \color{revblue}
  \chi_i(\mb{x}, t) = \frac{\epsilon a_i}{N-1} \nabla \textstyle\sum_{j\ne i}^{N}{a_j}.
  \color{black}
\end{equation}
%
\mpfive{This term contributes to the \emph{flux} of axonal branching as an
  additional source of diffusion, scaled by $\epsilon$, which reduces the
  branching density for a given projection where the branches of other
  projections are dense. Note that this operation is local to individual
  afferent projections.}

\MPone{In addition, the model we have outlined requires that molecular
  guidance gradients in the cortex are complemented by graded values of the
  interaction strengths, $\gamma$, at the level of the thalamus. While the
  precise mechanisms by which thalamic and cortical gradients interact during
  development have not been fully characterised, the presence of complementary
  thalamic and cortical molecular guidance gradients has been well established
  experimentally. In particular, the EphA4 receptor and its ligand ephrin-A5
  are distributed in complementary gradients in the somatosensory thalamus and
  cortex} (\citealp{vanderhaeghen_mapping_2000,miller_epha7-ephrin-a5_2006}).
\MPone{Cells originating in the medial part of ventrobasal thalamus
  (comprising VPM and VPL) express high levels of EphA receptors and project
  to the lateral part of S1, which expresses low levels of ephrin-A5, and
  cells originating in the lateral part of ventrobasal thalamus express low
  levels of EphA receptors and project to the medial part of S1, which
  expresses high levels of ephrin-A5} (see
\citealp{gao_regulation_1998,dufour_area_2003,vanderhaeghen_developmental_2004,speer_grading_2005,torii_role_2013}).
\MPone{We assume that such patterning arises because relative strengths of
  interaction with guidance molecules (e.g., ephrin-A5) in the cortex are
  correlated with the relative concentrations of complementary molecules
  (e.g., EphA4) in the thalamus, and thus with thalamic position along the
  axis to which their gradients are aligned.}

\MPone{For simplicity, the two simulated thalamic interaction gradients, as
  well as the two cortical guidance gradients, were initially chosen to be
  linear and orthogonal. Hence a given pair of $\gamma$ values corresponds to
  the coordinate of a barreloid center in the VPM. Coordinates, in a reference
  plane defined by the anterior-posterior and medial-lateral axes, were
  estimated from Fig 5d of} \citep{haidarliu_size_2001}, \MPone{and scaled
  such that $\gamma\in\pm 2$. Note that this scaling is arbitrary because
  according to the model the coordinates provide relative position information
  only.}

A cortical boundary enclosing \cmnt{the} 41 \cmnt{macrovibrissae} barrels was
traced from a cytochrome oxidase stain from \cite{zheng_signal_2001}
\cmnt{(using original data kindly supplied by the authors)}, and
Eqs.~\ref{eq:dc}--\ref{eq:comp} were solved for $N=41$ projections on the
resulting domain, $S$, using $M=2$ linear signalling gradients aligned with
the anterior-posterior and medial-lateral axes.  \mpfour{These gradients are
  shown with the barrel field boundary in Fig.\,\ref{fig:main}A for clarity,
  though like ephrin-5 they may be thought of as extending across the cortical
  hemisphere \citep{miller_epha7-ephrin-a5_2006}. Simulations were stepped
  through 30000 iterations of Eqs.\,\ref{eq:dc}--\ref{eq:comp} ($\delta
  t=0.0001$)}

\cmnt{Across a wide range of parameter values, random initial conditions (a
  uniform random distribution for $a(\mb{x},0)\in(0.2,0.4)$, $c(\mb{x},0)=0$)
  eventually yielded a clear Voronoi-like tessellation of topographically
  organised thalamocortical projections, confirming that barrel maps can
  self-organise in the absence of pre-specified center points. The
  organization is apparent in a plot of the identity of the projection for
  which the connection density is maximal at each simulated cortical location,
  as shown in Fig.\,\ref{fig:main}C. Parameters for the example simulation
  shown in Fig.\,\ref{fig:main}C (see also Movie S1) were obtained by
  conducting a full parameter sweep and choosing a combination of parameter
  values ($\alpha=3.6$, $\beta=16.67$, $k=3$, $D=0.5$, $\epsilon=1.2$) that
  scored well against the following three measures.}

% New section to describe and introduce the metrics
\metrics{First, we used an algorithm introduced by Honda to measure the
  discrepancy of each barrel shape from a Dirichletform shape}
\citep{honda_geometrical_1983}.  \metrics{Low overall values of this
  \emph{Honda-}$\delta$ metric obtained from simulated barrels indicate a
  close correspondence of the simulated barrel field with a Voronoi
  tesselation, and thus with a biological barrel field. For the tesselation
  that is overlaid on the real barrel field in Fig.\,\ref{fig:main}A,
  $\delta=0.025$, and a reduction in $\delta$ in the example simulation over
  time confirmed that an equivalent `good' Voronoi pattern} (see
\citealp{senft_mouse_1991}) \metrics{can emerge within $\approx$~20000
  iterations (Fig.\,\ref{fig:main}D, red circles). Second, we devised a
  \emph{pattern difference} measure that is sensitive to deviations in the
  component shapes and overall topographic registration between two
  tesselations, $\eta$, and we used this measure to compare the simulated
  barrel fields to the real barrel field from which the boundary shape applied
  to the simulation was obtained (see \emph{Materials and Methods} for
  details). A similar reduction in $\eta$ in the development of the example
  simulation confirmed that the shapes and arrangement of emergent connection
  fields came to match those of the real barrel field by around 20000
  iterations (Fig.\,\ref{fig:main}D, black squares). Third, we measured the
  \emph{connection selectivity} at each location on the cortical sheet, as the
  connection density of the most dense projection divided by the sum over all
  projection densities, $\omega$. The connection selectivity increased as the
  barrel map self-organised in the example simulation (Fig.\,\ref{fig:main}D,
  grey hexagons), and the selectivity became concentrated in regions
  overlapping with the emergent barrel centers (Fig.\,\ref{fig:main}E).}

\metrics{Against these three metrics we are also able to characterise the
  robustness of self-organization to the model parameters, and to investigate
  the sensitivity of the model to variation in its inputs.}

@@

\MPthreePar{Fig.\,\ref{fig:paramsweep}A shows values of $\delta$, $\eta$ and
  $\omega$ obtained after 30000 iterations, from 216 independent simulations,
  each representing a unique combination of the model parameters $D$,
  $\epsilon$, and the ratio $\alpha/\beta$. First we observe that
  self-organization is highly robust to the ratio $\alpha/\beta$, across five
  orders of magnitude, with respect to all three metrics. Second, the most
  strongly Dirichletform patterns (low \emph{Honda}-$\delta$) were generated
  by simulations in which the diffusion constant $D$ and the strength of
  competition $\epsilon$ were high. Third, strongest connection selectivities,
  $\omega$, were obtained for lower values of $D$. Fourth, variation in the
  pattern difference metric, $\eta$, indicated that the alignment between real
  and simulated patterns was greatest for intermediate rates of diffusion,
  $D\approx0.5$. Together these results indicate that when competition is
  strong, the rate of diffusion determines a trade-off such that fields emerge
  to be barrel-shaped when diffusion is fast and they emerge to be more
  selectively innervated when diffusion is slow.}

\MPthreePar{The parameters of the example simulation are indicated in
  Fig.\,\ref{fig:paramsweep}A using an asterisk. In
  Fig.\,\ref{fig:paramsweep}B, we also present examples of alternative
  patterns that emerge for different choices of $D$. Decreasing the rate of
  diffusion may be considered equivalent to increasing the overall size of the
  domain, $S$, hence insights into barrel development in species with
  substantially larger barrel fields, such as capybara which do not have
  clearly identifiable barrels, may be gained by studying pattern formation
  when $D$ is small. In this context, it is interesting to note that for small
  $D$, the organization is predicted to be topological but highly irregular,
  with a general expansion in the territory occupied by the central versus
  peripheral domains that would presumably manifest as an absence of
  identifiable barrel fields (Fig.\,\ref{fig:paramsweep}Biii).}

@@

\MPthreeSens{Next we conducted a sensitivity analysis to determine the extent
  to which the quality of the pattern (after $t=30000$ iterations) is affected
  by perturbations to i)~the magnitude and offset of the noise applied to
  $a_i$ at $t=0$; ii)~noise applied to the interaction parameters,
  $\gamma_{i,j}$; iii)~noise (at various length scales) applied to the
  guidance fields; and iv)~the magnitude and orientation of one cortical
  guidance field relative to the other (Fig.\,\ref{fig:main}A).}

\MPthreeSens{Using the parameters of the example simulation
  (Fig.\,\ref{fig:main}C) we established baseline mean and standard deviations
  from ten independent simulations with initial uniform random values for
  $a(\mb{x},0)\in(0.2,0.4)$, to be $\delta=0.089\pm 0.004$, $\eta=0.2108\pm
  0.002$, and $\omega=0.2165\pm 0.0001$. Repeating with the variation in the
  initial noise doubled ($a(\mb{x},0)\in(0.1,0.5)$), or removed altogether
  ($a(\mb{x},0)=0.3$), generated distributions of $\delta$, $\eta$, and
  $\omega$ that were not statistically different, as established using paired
  two-sample t-tests. Adding noise to the interaction parameters ($\gamma$)
  affected neither the \emph{Honda-}$delta$ or the \emph{connection
    selectivity} measures substantially (see Fig\,\ref{fig:sens}A), and an
  increase in the \emph{pattern difference} reflected an increase in the
  occurrence of topological defects only when perturbations become so large as
  to cause the ordering of $\gamma$ values from neighbouring thalamic sites to
  be switched (see example map Fig\,\ref{fig:sens}Bi). Adding noise to the
  cortical guidance field values, $\rho_1(\mathbf{x})$ and
  $\rho_2(\mathbf{x})$, disrupted pattern formation only for high levels of
  noise applied at short length scales, which manifested as non-straight edges
  at the domain boundaries (Fig\,\ref{fig:sens}Bii). Varying the slope of one
  linear gradient $\rho_1(\mathbf{x})$ while keeping that of the other
  constant caused elongation of the emergent domains along the corresponding
  axis (Fig\,\ref{fig:sens}Biii), while pattern formation was not strongly
  influenced by relaxing the assumption that the gradients of the two cortical
  guidance fields are orthogonal (Fig\,\ref{fig:sens}Biv). Overall, the
  sensitivity analysis revealed that self-organization of barrel-like fields
  in the model is highly robust to a wide range of sources of perturbation.}

% This should be a little less terse than in the original.
\MPthreePred{To further investigate the interplay of genetic and intrinsic
  mechanisms we simulated two well known experimental manipulations of the
  whisker barrel system. First, we simulated a seminal barrel duplication
  experimental paradigm}
\citep{shimogori_fibroblast_2005,assimacopoulos_fibroblast_2012}
\MPthreePred{in which the growth factor Fgf8, which is normally expressed at
  the anterior end of the cortical subplate from around E9.5}
\citep{crossley_mouse_1995}, \MPthreePred{is ectopically expressed (by
  electroporation) also at the posterior pole of a cortical
  hemisphere. Because Fgf8 sets off a chain of gene interactions involved in
  controlling the graded expression of ephrin and cadherin molecules, its
  misexpression at the posterior pole could be expected to mirror those
  gradients which usually vary along the rostrocaudal axis. We assume that
  this results in a mirror of the primary barrel cortex boundary along the
  rostrocaudal axis, to match the experimental results of}
\citealp{assimacopoulos_fibroblast_2012}. \MPthreePred{We also assume that a
  further consequence of the posterior Fgf8 electroporation is the mirrored
  anterior-posterior guidance gradient $\rho_1$ (Fig.\,\ref{fig:fgf8}A). With
  the modified $\rho_1$ guidance gradient the simulation was run with $N=41$
  thalamocortical axon types, using the parameters of the example
  simulation. The result, after 30000 iterations, was two mirror-symmetrical
  barrel fields comprising $2N$ barrels (Fig\,\ref{fig:main}B), consistent
  with the outcome of the original experiments.}

@@

\MPthreePred{To investigate the response of the model to environmental
  manipulation, we simulated a whisker deprivation experiment. In the first
  few days after birth, the upper region of the rat's cortex develops into
  distinct layers I--IV (layers V and VI are already distinct at
  birth). During this critical period, removal of the mystacial whiskers by
  electrocauterization, plucking or trimming leads to observable changes in
  brain structures, including the barrel field in cortical layer
  IV~\citep{jeanmonod_mouse_1981}. Amoungst other changes, barrels associated
  with deprived whiskers have a smaller area. This is found whether the animal
  is deprived of an individual whisker~\citep{kossut_effects_1992}, or an
  entire row~\citep{loos_somatosensory_1973}, and demonstrates that
  thalamocortical activity contributes to the developmental process. We
  simulated the trimming of the individual whisker C3 during this critical
  period by reducing the amount by which its thalamocortical projections
  compete for branching. This was achieved by reducing the value of $\epsilon$
  in Eq.\,\ref{eq:comp} for the C3 projection, leaving other simulation
  parameters unchanged. The reduced competition for branching limited the area
  over which the C3 projection could make connections and its resulting field
  size was diminished (Fig.\,\ref{fig:fgf8}C). Emergent fields representing
  the neighbours of the C3 projection, whose ability to compete for branching
  was unchanged, increased in size. The C3 barrel disappeared altogether if
  its competition parameter was reduced below a half of the original value. A
  reduction in the area of barrel C3, comparable that induced by
  \citep{kossut_effects_1992} (65\%) was obtained in simulation when the
  competitiveness of projection C3, $\epsilon$, was reduced to 86\%, as
  indicated in (Fig.\,\ref{fig:fgf8}D).}

\MPthreePred{Together with the results of the misexpression experiment, the
  consistency of the simulated whisker trimming results with those of the
  original studies demonstrates how the model can be used to test specific
  hypotheses about the contribution of intrinsic and extrinsic factors to the
  development of cortical fields.}

@@

% for ``end of methods''
\MPone{We concentrated on the representation of the forty-six macrovibrissae
  that constitute a given barrel field, because their thalamic and cortical
  correlates are easily identifiable and consistently positioned. We exclude
  the representation of the microvibrissae to limit the overall complexity of
  the simulations, as well as the five rhinal whiskers as their representation
  in S1 is isolated from the main barrel field. Consequently, we set $N=41$.}
